---
title: "SDM with R"
author: "Pedro Henrique Braga and Julia Nordlund"
date: "April 23, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Outline

This tutorial illustrates how to build, evaluate and project species
distribution models within R.

The main steps, described bellow, are the following:

1. Data preparation
    + Study area and manipulating spatial information
    + Environmental data
    + Species occurrence data
        + Presence-and-absence data
        + Presence-only and pseudo-absences data
2. Model fitting, prediction, and evaluation
3. Making projections

# Workshop environment preparation

Install and load all required *R* libraries:

```{r, eval=FALSE}
install.packages("rgdal")
install.packages("raster")
install.packages("rgeos")
install.packages("dismo")
install.packages("letsR")
install.packages("biomod2")
install.packages("biogeo")
install.packages("maptools")
```

```{r, message=FALSE, warning=FALSE}
library(rgdal)
library(raster)
library(rgeos)
library(dismo)
library(letsR)
library(biomod2)
library(biogeo)
library(maptools)
```

Download the .zip file from xxxx and extract files to your QCBS R Symposium workshop folder.

Then, set your working directory to the same directory of the folder `workshopSDM` using the functions `setwd()` or your prefered method.

# Data preparation

## Selecting your study area and manipulating spatial files

Most species distribution models are done using spatial grid information. In the next step, we will learn how to create a polygon grid of a given region, which will be used within our species distribution model framework.

For this tutorial, we have chosen the Neotropical zoogeographical region as our study area. Load the polygon shapefile of it using the `readOGR` function:

```{r, echo = TRUE}
neotropical_shape <- rgdal::readOGR("data/shapefiles/neotropical.shp")
plot(neotropical_shape)
```

Now that we have our study region, we need to create a grid, intersect our shapefile and create a new grid of that region. For this, we will use the `GridFilter` function. It allows us to input a polygon shapefile, specify a resolution and the proportion of overlap for each cell to be considered.

We will apply it , intersect and crop its features with the polygon `neotropical_shape` you have loaded.

```{r, eval = FALSE}
GridFilter <- function(shape, resol = 1, prop = 0)
{

  grid <- raster(extent(shape))
  res(grid) <- resol
  proj4string(grid)<-proj4string(shape)
  gridpolygon <- rasterToPolygons(grid)
  drylandproj<-spTransform(shape, CRS("+proj=laea"))
  gridpolproj<-spTransform(gridpolygon, CRS("+proj=laea"))
  gridpolproj$layer <- c(1:length(gridpolproj$layer))
  areagrid <- gArea(gridpolproj, byid=T)
  gridpolproj <- gBuffer(gridpolproj, byid=TRUE, width=0)
  dry.grid <- intersect(drylandproj, gridpolproj)
  areadrygrid <- gArea(dry.grid, byid=T)
  info <- cbind(dry.grid$layer, areagrid[dry.grid$layer], areadrygrid)
  dry.grid$layer<-info[,3]/info[,2]
  dry.grid <- spTransform(dry.grid, CRS(proj4string(shape)))
  dry.grid.filtered <- dry.grid[dry.grid$layer >= prop,]
}

# Create a spatial polygon grid for the Neotropical region, with 5 degrees  x 5 degrees
neotropical_grid <- GridFilter(neotropical_shape, resol = 5, prop = 0.5)


```

```{r, eval = FALSE}
# Export your resulting polygon grid 
writeOGR(neotropical_grid, dsn=paste(getwd(),"/data/shapefiles", sep=""), layer="neotropical_grid_5", driver="ESRI Shapefile", overwrite_layer=T)
```

This is our resulting polygon grid, where each cell refers to a site (and a row) in our data.

```{r raster, echo=TRUE, warning=FALSE}
neotropical_grid <-  shapefile("data/shapefiles/neotropical_grid_5.shp")
plot(neotropical_grid)

```


```{r, echo = FALSE}
# extract coordinates from the cell centroids
coords <- as.data.frame(coordinates(neotropical_grid))
colnames(coords) <- c("Longitude_X", "Latitude_Y")

write.table(coords, "data/matrices/NT_coords_5.txt")
```


## Importing environmental variables



```{r, eval = FALSE}


bio_1 <- raster("/data/rasters/w25m_bio_1.asc")

myExpl <- data.frame(bio_1 = numeric(length(neotropical_grid))
                     )

bio_1_ext <- extract(bio_1,neotropical_grid)

myExpl$bio_1 <- unlist(lapply(bio_1_ext, 
                              function(x) 
                                if (!is.null(x)) mean(x, na.rm=TRUE) 
                              else NA))
# Let's take a look at our table:

head(myExpl)

write.table(myExpl, "data/matrices/NT_EnvVar_5.txt")

```



## Importing species data
### Using expert drawn maps

```{r maptools, echo=TRUE, warning=FALSE}
speciesGeoDist <- readShapePoly("data/shapefiles/panthera_onca_IUCN.shp",  
                                   proj4string=CRS("+proj=laea"))
plot(speciesGeoDist)

```


```{r, echo=FALSE, warning=FALSE}
neotropical_grid <- readShapePoly("data/shapefiles/neotropical_grid_5.shp",  
                                   proj4string=CRS("+proj=laea"))


```


Create a presence-absence matrix of species' geographic ranges within your polygon grid shapefile:
```{r, echo = FALSE}
abspres.matrix <- lets.presab.grid(speciesGeoDist, neotropical_grid, "ID")
```


You can now visualize a map of your species richness within your polygon grid:
```{r, echo = FALSE}
richnessCalc <- rowSums(abspres.matrix$PAM) + 1
colPalette <- colorRampPalette(c("#fff5f0", "#fb6a4a", "#67000d"))
colors <- c("white", colPalette(max(richnessCalc)))
plot(abspres.matrix$grid, border = "gray40",
     col = colors[richnessCalc])
map(add = TRUE)
```

Export your presence-absence matrix:
```{r, echo = FALSE}
write.table(abspres.matrix$PAM, "data/matrices/NT_PAM_5.txt")
```


### Using occurrence records
## Data cleaning

```{r, echo = FALSE}
# Load our species data
DataSpecies <- read.table("data/matrices/NT_PAM_5.txt", header=TRUE)
head(DataSpecies)

# Remove all columns that have less than 4 presences
to.remove <- colnames(DataSpecies)[colSums(DataSpecies) <= 4]
`%ni%` <- Negate(`%in%`)
DataSpecies <- subset(DataSpecies,select = names(DataSpecies) %ni% to.remove)

# Replace "_" per "." 
names(DataSpecies) <- gsub(x = names(DataSpecies),
                           pattern = "\\.",
                           replacement = ".")

DataSpecies <- read.table("data/matrices/NT_PAM_5.txt", header=TRUE)

```

## Species distribution modelling

First, input data needs to be rearranged for usage in biomod2 using `BIOMOD_FormatingData()`. Its most important arguments are: 
* `resp.var` contains **species data** (for a single species) in **binary format** (ones for presences, zeros for true absences and NA for indeterminated) that will be used **to build the species distribution models**. It may be a vector or a spatial points object.
* ``expl.var`` contains a matrix, data.frame, SpatialPointsDataFrame or RasterStack containing
your explanatory variables.
* ``resp.xy`` two columns matrix containing the X and Y coordinates of resp.var (only consider if resp.var is a vector)
* ``resp.name`` contains the response variable name (species).


```{r, echo = TRUE}

# Recall your variables again: 

## Load prepared species data
myResp <-  read.table("data/matrices/NT_PAM_5.txt", header=TRUE)

## Load environmental variables
myExpl <-  read.table("data/matrices/NT_EnvVar_5.txt", header=TRUE)

# Define the species of interest
myRespName <- names(DataSpecies)[1]

# Load coordinates
myRespCoord <- read.table("data/matrices/NT_coords_5.txt", header=TRUE)

myBiomodData <- BIOMOD_FormatingData(resp.var = myResp,
                                     expl.var = myExpl,
                                     resp.xy = myRespCoord,
                                     resp.name = myRespName,
                                     na.rm = TRUE)
```

```{r, eval = FALSE}
myBiomodOption <- BIOMOD_ModelingOptions()
```


```{r, eval = FALSE}
myBiomodModelOut <- BIOMOD_Modeling(
    myBiomodData,
    models = c('MAXENT.Tsuruoka', 'RF'),
    models.options = myBiomodOption,
    NbRunEval=3,
    DataSplit=80,
    Prevalence=0.5,
    VarImport=3,
    models.eval.meth = c('TSS','ROC'),
    SaveObj = TRUE,
    rescal.all.models = FALSE,
    do.full.models = FALSE,
    modeling.id = paste(myRespName,"CurrentClim",sep=""))
```


```{r, eval = FALSE}
  ### save models evaluation scores and variables importance on hard drive
  capture.output(get_evaluations(myBiomodModelOut),
                 file=file.path(myRespName,
                                paste(myRespName,"_formal_models_eval.txt", sep="")))
  capture.output(get_variables_importance(myBiomodModelOut),
                 file=file.path(myRespName,
                                paste(myRespName,"_formal_models_var_imp.txt", sep="")))
```


```{r, eval = FALSE}
  ### Building ensemble-models
  myBiomodEM <- BIOMOD_EnsembleModeling(
    modeling.output = myBiomodModelOut,
    chosen.models = 'all',
    em.by='all',
    eval.metric = c('TSS'),
    eval.metric.quality.threshold = c(0.5),
    prob.mean = T,
    prob.cv = T,
    prob.ci = T,
    prob.ci.alpha = 0.05,
    prob.median = T,
    committee.averaging = T,
    prob.mean.weight = T,
    prob.mean.weight.decay = 'proportional')
```

```{r, eval = FALSE}
  ### Make projections on current variable
  myBiomodProj <- BIOMOD_Projection(
    modeling.output = myBiomodModelOut,
    new.env = myExpl,
    xy.new.env = myRespCoord,
    proj.name = 'current',
    selected.models = 'all',
    binary.meth = 'TSS',
    compress = TRUE,
    clamping.mask = F,
    output.format = '.RData')


myCurrentProj <- getProjection(myBiomodProj)
myCurrentProj

```

  
```{r, eval = FALSE}
  ### Make ensemble-models projections on current variable
  myBiomodEF <- BIOMOD_EnsembleForecasting(
    projection.output = myBiomodProj,
    binary.meth = 'TSS',
    total.consensus = TRUE,
    EM.output = myBiomodEM)
```

# print summary
myBiomodEM

# get evaluation scores
getEMeval(myBiomodEM)

```{r, eval = FALSE}
  ### Make ensemble-models projections on current variable
# load the first speces binary maps which will define the mask

alphaMap <- reclassify(subset(myExpl,1), c(-Inf,Inf,0))

alphaMap <- get(load(paste(getwd(),"/",myRespName[1],"/proj_current/", "proj_current_",
                           myRespName[1],"_ensemble_TSSbin.RData",
                           sep="")))[,1]

``` 